{\bf {\Large Improving estimates of ecosystem metabolism by reducing effects of tidal advection on dissolved oxygen time series}}

### 2015-02-03 abstract.tex
cover letter, some edits from Jim, waiting on final comments before first submissions

In aquatic ecosystems, time series of \ac{DO} have been used to compute estimates of integrated ecosystem metabolism.  Central to this open water or ``Odum'' method is the assumption that the dissolved oxygen time series is a Lagrangian specification of the flow field.  However, most \ac{DO} time series are collected at fixed locations, such that the method must assume changes in dissolved oxygen principally reflect ecosystem metabolism and that effects due to advection or mixing can be neglected.  A statistical model using weighted regression was applied to separate variability in \ac{DO} associated with metabolism from tidal variation or other advection in estuaries, thereby helping to partially relax this assumption and improve estimates of ecosystem metabolism. The method was developed and tested using a simulated \ac{DO} time series with known biological and physical components, and then applied to one year of continuous monitoring data from four water quality stations within the \acl*{NERRS}.  Overall, the approach is a useful way to reduce variability in estimates of ecosystem metabolism caused by advection, particularly when the magnitude of tidal influence is high and correlations between tidal change and solar cycling are low. By reducing the effects of physical transport on metabolism estimates, there may be increased potential to empirically relate metabolic rates to causal factors on times scales of several days to several weeks. Estimates of variability associated with physical advection may also be more interpretable, since convolution of physical and biological effects can be reduced.

>>>>

In aquatic ecosystems, time series of \ac{DO} have been used to compute estimates of integrated ecosystem metabolism.  Central to this open water or ``Odum'' method is the assumption that the dissolved oxygen time series is a Lagrangian specification of the flow field.  However, most \ac{DO} time series are collected at fixed locations, such that the method must assume changes in dissolved oxygen principally reflect ecosystem metabolism and that effects due to advection or mixing can be neglected.  A statistical model using weighted regression was applied to separate variability in \ac{DO} associated with metabolism from tidal variation or other advection in estuaries, thereby helping to partially relax this assumption and improve estimates of ecosystem metabolism. The method was developed and tested using a simulated \ac{DO} time series with known biological and physical components, and then applied to one year of continuous monitoring data from four water quality stations within the \acl*{NERRS}.  Overall, the approach is a useful way to reduce variability in estimates of ecosystem metabolism caused by advection, particularly when the magnitude of tidal influence is high and correlations between tidal change and the solar cycle are low. By reducing the effects of physical transport on metabolism estimates, there may be increased potential to empirically relate metabolic rates to causal factors on times scales of several days to several weeks. Estimates of variability associated with physical advection may also be more interpretable, since convolution of physical and biological effects can be reduced.

### 2015-02-03 intro.Rnw
cover letter, some edits from Jim, waiting on final comments before first submissions

Ecosystem metabolism describes the balance between production and respiration processes that create and consume organic matter \citep{Kemp12,Needoba12}.  Light exposure experiments of water samples collected at discrete locations and times have traditionally been used to measure metabolic activity.  Although highly controlled and precise, bottle-based methods are labor-intensive and not scalable to describe entire ecosystem rates.  Bottle-based methods may also only reliably estimate production and respiration associated with planktonic processes, whereas significant contributions of ecosystem production can arise from other habitats, such as benthic seagrass communities.  By contrast, open-water techniques have been increasingly used to estimate whole system metabolism given the availability of long-term, continuous time series of dissolved oxygen \citep{Odum56,Davanzo96}. Daily integrated measurements of metabolism represent the balance between daytime production and nighttime respiration attributed to all ecosystem components.  Open-water estimates also provide a basis for tracking ecosystem change over time and are more practical for capturing events or evaluating trends as compared to bottle-based approaches. Although metabolic rates vary naturally at different spatial and temporal scales \citep{Ziegler98,Caffrey04,Russell07}, anthropogenic nutrient sources are often contributing factors that increase rates of production (\citealt{Nixon95}, \citetalias{NRC00}).  Reliables estimates of whole ecosystem metabolism are critical for measuring both background rates of production and potential impacts of human activities on ecosytem condition.

The ability to accurately estimate whole system metabolism using the open-water method depends on the degree to which assumptions of the theory are met \citep{Staehr10,Kemp12}.  The fundamental assumption is that the time series of \ac{DO} represents a Lagrangian specification of the flow field that describes the same water mass over time \citep{Needoba12}.  The Lagrangian specification assumes that the time series characterizes individual fluid parcels regardless of location, as in a parcel of water moving with the tide.  In reality, most \ac{DO} time series are collected at fixed locations such as a mooring or dock, which is characteristized by an Eulerian specification of the flow field.  Time series at fixed locations may characterize water masses with different metabolic histories if water particles are transported by physical advection. A Lagrangian flow field is often assumed, such that estimates of metabolism may be inaccurate if substantial variation in water column mixing occurs throughout the period of observation \citep{Kemp80,Russell07}.  Given this critical challenge, the open-water method has been used with varying success in estuaries influenced by tidal mixing \citep{Caffrey04,Russell07,Caffrey14}.  In contrast, the method has been more successfully applied to water quality time series in lakes, although stratification may limit estimates to specific vertical layers \citep{Staehr10,Coloso11,Batt12}. Appropriate placement of monitoring sondes, sampling frequency and duration, and reliability of data from single stations have been relevant issues in applying the open-water method to systems influenced by physical mixing \citep{Russell07,Staehr10}.  Individual sampling stations near bay inlets or along major tidal axes may produce \ac{DO} time series that fail to meet the assumptions of the open-water method.

Although numerous studies have shown that application of the open-water method to lakes or estuaries may be problematic \citep{Ziegler98,Caffrey03,Coloso11,Batt12,Nidzieko14}, very few quantitative approaches have been developed to address potential bias or noise in \ac{DO} signals from physical advection.  For example, an extensive analysis by \citet{Caffrey03} applied the open-water method to estimate metabolism at 28 continuous monitoring stations at 14 US estuaries.  A significant portion of the production and respiration estimates were negative (3 - 69\% depending on site), suggesting advection of water masses was a likely factor influencing the \ac{DO} time series.  These `anomalous' values are typically omitted from the analysis \citep{Caffrey03,Collins13}, which may upwardly bias estimates of metabolism \citep{Caffrey14}.  Further, \citet{Nidzieko14} evaluated the effects of tidal advection on metabolism estimates in a mesotidal estuary.  Estimates from a single location were strongly correlated with the spring-neap cycle such that net heterotrophy was more common during spring tides, whereas metabolism was generally balanced during neap tides.  A control-volume approach was used by impounding a section of the upper estuary to understand how physical processes contribute to biological variability.  Although useful as an \textit{in situ}, site-specific approach, more accessible statistical methods specific to time series are needed given the increasing availability of continuous monitoring data. For example, \citet{Batt12} explored the use of a Kalman filter \citep{Harvey89} to remove process and observation uncertainty from \ac{DO} time series in lakes.  Similar approaches have not been developed for estuaries, particularly those that address potential effects of tidal advection.

This article describes the development and application of a method for improving estimates of ecosytem metabolism computed from \ac{DO} time series.  Specifically, the apparent effects of tidal advection on \ac{DO} observations are removed to improve the fidelity of open-water metabolism estimates derived from continuous water quality data.  We used a weighted regression approach originally developed to resolve trends in pollutant concentrations in streams and rivers \citep{Hirsch10}.  The weighted regression approach creates dynamic predictions of \ac{DO} as a function of time and tidal height change, which are then used to filter, or detide, the \ac{DO} signal.  The model is based on the recognition that daily fluctuations in \ac{DO} are caused by metabolism associated with the solar cycle, whereas other fluctuations in estuaries are likely associated with water level changes that generally exhibit pregression relative to the solar cycle.  The weighted regression model was applied, rather than methods commonly used for detiding in physical oceanography, to allow for the complex and dynamic patterns of \ac{DO} changes relative to advection.  First, we used simulated \ac{DO} time series with known characteristics to evaluate ability of the weighted regression to remove the simulated effects of a tidally-advected \ac{DO} gradient.  Second, the simulation results informed the application of the method to four case studies chosen from the \aclu{NERRS} (\acs{NERRS}, \citealt{Wenner04}).  In all examples, tidal height is used as a proxy for lateral water movements that may influence \ac{DO} observations.  In the absence of quantitative data describing lateral \ac{DO} variation (e.g., contemporaneous stations along a tidal axis), we assume tidal height is an appropriate characterization of lateral variation.  Accordingly, `tidal variation' or `changes in tidal height' are used throughout in reference to assumed lateral \ac{DO} gradients that are carried past monitoring sensors by tidal currents.

>>>>

Ecosystem metabolism describes the balance between production and respiration processes that create and consume organic matter \citep{Kemp12,Needoba12}.  Light exposure experiments of water samples collected at discrete locations and times have traditionally been used to measure metabolic activity.  Although highly controlled and precise, bottle-based methods are labor-intensive and not scalable to describe entire ecosystem rates.  Bottle-based methods may also only reliably estimate production and respiration associated with planktonic processes, whereas significant contributions to ecosystem metabolism can arise from other habitats, such as the benthos or intertidal marshes.  By contrast, open-water techniques have been increasingly used to estimate whole system metabolism given the availability of long-term, continuous time series of dissolved oxygen \citep{Odum56,Davanzo96}. Daily integrated measurements of metabolism represent the balance between daytime production and nighttime respiration attributed to all ecosystem components.  Open-water estimates also provide a basis for tracking ecosystem change over time and are more effective for capturing events or evaluating trends as compared to bottle-based approaches. Although metabolic rates vary naturally at different spatial and temporal scales \citep{Ziegler98,Caffrey04,Russell07}, anthropogenic nutrient sources are often contributing factors that increase rates of production (\citealt{Nixon95}, \citetalias{NRC00}).  Reliables estimates of whole ecosystem metabolism are critical for measuring both background rates of production and potential impacts of human activities on ecosytem function.

The ability to accurately estimate whole system metabolism using the open-water method depends on the degree to which assumptions of the theory are satisfied \citep{Staehr10,Kemp12}.  The fundamental assumption is that the time series of \ac{DO} represents a Lagrangian specification of the flow field \citep{Needoba12}.  The Lagrangian specification characterizes an individual fluid parcel through time, such as a parcel of water moving with the tide.  In reality, most \ac{DO} time series are collected at fixed locations such as a mooring or dock, so they are a Eulerian specification of the flow field.  Time series at fixed locations may characterize water masses with different metabolic histories as water is advected past the sensor, leading to errors in metabolism estimates \citep{Kemp80,Russell07}.  Given this critical challenge, the open-water method has been used with varying success in estuaries influenced by tidal transport and mixing \citep{Caffrey04,Russell07,Caffrey14}.  In contrast, the method has been more successfully applied to water quality time series in lakes, although stratification may limit estimates to specific vertical layers \citep{Staehr10,Coloso11,Batt12}. Appropriate placement of monitoring sondes, sampling frequency and duration, and reliability of data from single stations have been relevant issues in applying the open-water method to systems influenced by physical mixing \citep{Russell07,Staehr10}.  Individual sampling stations near bay inlets or along major tidal axes may produce \ac{DO} time series that are strongly influenced by advection leading to relatively large errors when using the open-water method.

Although numerous studies have shown that application of the open-water method to lakes or estuaries may be problematic \citep{Ziegler98,Caffrey03,Coloso11,Batt12,Nidzieko14}, very few quantitative approaches have been developed to address potential bias or noise in metabolism estimates resulting from advection.  For example, an extensive analysis by \citet{Caffrey03} applied the open-water method to estimate metabolism at 28 continuous monitoring stations at 14 US estuaries.  A significant portion of the production and respiration estimates were negative (3 - 69\% depending on site), suggesting advection of water masses was a likely a major factor influencing the \ac{DO} time series.  These `anomalous' values are typically omitted from the analysis \citep{Caffrey03,Collins13}, which may upwardly bias estimates of mean metabolism \citep{Caffrey14}.  Further, \citet{Nidzieko14} evaluated the effects of tidal advection on metabolism estimates in a mesotidal estuary.  Estimates from a single location were strongly correlated with the spring-neap cycle such that net heterotrophy was more common during spring tides, whereas metabolism was generally balanced during neap tides.  This examples illustrates that metabolism may actually be modulated by the tide, as opposed to artifacts in the \ac{DO} time series caused by advection on daily time scales.  \citet{Nidzieko14} used a control-volume approach by impounding a section of the upper estuary to understand how physical processes contribute to biological variability.  Although useful as an \textit{in situ}, site-specific approach, more accessible statistical methods specific to time series are needed given the increasing availability of continuous monitoring data. For example, \citet{Batt12} explored the use of a Kalman filter \citep{Harvey89} to remove process and observation uncertainty from \ac{DO} time series in lakes.  Similar approaches have not been developed for estuaries, particularly those that address potential effects of tidal advection.

This article describes the development and application of a method for improving estimates of ecosytem metabolism computed from \ac{DO} time series by reducing the effect of tidal advection on continuous \ac{DO} observations.  We used a weighted regression approach originally developed to resolve trends in pollutant concentrations in streams and rivers \citep{Hirsch10}.  The weighted regression approach creates dynamic predictions of \ac{DO} as a function of time and tidal height change, which are then used to filter, or detide, the \ac{DO} signal.  The model is based on the recognition that daily fluctuations in \ac{DO} caused by metabolism are associated with the solar cycle, whereas other fluctuations in estuaries are likely associated with cyclical water level changes that generally exhibit pregression relative to the solar cycle.  The weighted regression model was applied, rather than methods commonly used for detiding in physical oceanography, to allow for the complex and dynamic patterns of \ac{DO} changes relative to advection.  During initial method development, we used simulated \ac{DO} time series with known characteristics to evaluate ability of the weighted regression to remove the simulated effects of a tidally-advected \ac{DO} gradient.  Subsequently, the simulation results informed further development of the method as applied to four case studies chosen from the \aclu{NERRS} (\acs{NERRS}, \citealt{Wenner04}).  In all examples, tidal height is used as a proxy for advection because it was recorded with the \ac{NERRS} data.  In the absence of quantitative data describing lateral \ac{DO} variation (e.g., contemporaneous stations along a tidal axis), the model determines the advective effect empirically.

### 2015-02-03 mat_proc.Rnw
cover letter, some edits from Jim, waiting on final comments before first submissions

For this study, we adapted a weighted regression model to filter \ac{DO} time series for apparent tidal effects.  This model relied heavily on concepts used to develop the \ac{WRTDS} method for estimating pollutant concentrations in streams and rivers \citep{Hirsch10}.  The functional form of the model is:
\begin{equation}\label{funform}
DO_{obs}= \beta_0 + \beta_1 t + \beta_2 H
\end{equation}
where $DO_{obs}$ is a linear function of time $t$ and tidal height $H$.  Time is a continuous variable for the day and time of each observation as a proportion of the number of total observations added to each day.  The beginning of each day was defined as the nearest thirty minute observation to sunrise for the location.  Our model differed from the original \ac{WRTDS} method that included parameters to estimate variation of the response variable on a sinuisoidal period.  \ac{DO} variation was not modeled using this approach to avoid constraining parameter estimates by periodic, diel components.  Additionally, a more desirable parameter for modelling the effects of tidal advection on observed \ac{DO} would be a direct measure of horizontal water movement, rather than tidal height as used above.  Although this approach would more accurately describe the theoretical foundation of the mechanism, tidal height is an appropriate proxy given that the model is empirical and only requires a variable that is correlated with the true measure of interest.  We suspect that differences in results using a measure of lateral advection as compared to tidal height as a predictor would be minimal.

>>>>

For this study, we adapted a weighted regression model to filter \ac{DO} time series for apparent tidal effects.  This model relied heavily on concepts used to develop the \ac{WRTDS} method for estimating pollutant concentrations in streams and rivers \citep{Hirsch10}.  The functional form of the model is:
\begin{equation}\label{funform}
DO_{obs}= \beta_0 + \beta_1 t + \beta_2 H
\end{equation}
where $DO_{obs}$ is a linear function of time $t$ and tidal height $H$.  Time is a continuous variable for the day and time of each observation as a proportion of the number of total observations.  The beginning of each day was defined as the nearest thirty minute observation to sunrise for the location.  Our model differed from the original \ac{WRTDS} method, which included parameters to estimate variation of the response variable on a sinusoidal period.  \ac{DO} variation was not modeled using this approach because the expected metabolically-driven pattern is not strictly sinusoidal.  Tidal height is used as a proxy for advection because the model is empirical and only requires a variable that is correlated with the true measure of interest.  Moreover, tidal height measurements are more widely available and are generally monitored by water quality sondes.

### 2015-02-03 swmp_noise.Rnw
cover letter, some edits from Jim, waiting on final comments before first submissions

\raggedbottom
\linenumbers
\raggedright
\urlstyle{same}
\setlength{\parindent}{0.5in}
\renewcommand\refname{References \vspace{12pt}}

@@

We acknowledge the significant efforts of research staff and field crews from the \acl{SWMP} of the \acl{NERRS} for providing access to high quality data sets.  We thank Dr. Jane Caffrey for reviewing a manuscript draft and previous work on applications of the open-water method to estuarine monitoring data. This study was funded by the US \acl{EPA}, but the contents are solely the views of the authors.  Use of trade names does not constitute endorsement by the US government.

>>>>

\raggedbottom
% \linenumbers
\raggedright
\urlstyle{same}
\setlength{\parindent}{0.5in}
\renewcommand\refname{References \vspace{12pt}}

@@

We acknowledge the significant efforts of research staff and field crews from the \acl{SWMP} of the \acl{NERRS} for providing access to high quality data sets.  We thank Dr. Jane Caffrey and Dr. Erik Smith for reviewing early drafts of the manuscript. The views expressed in this article are those of the authors and do not necessarily reflect the views or policies of the U.S. Environmental Protection Agency.

### 2015-02-03 swmp_noise.tex
cover letter, some edits from Jim, waiting on final comments before first submissions

Ecosystem metabolism describes the balance between production and respiration processes that create and consume organic matter \citep{Kemp12,Needoba12}.  Light exposure experiments of water samples collected at discrete locations and times have traditionally been used to measure metabolic activity.  Although highly controlled and precise, bottle-based methods are labor-intensive and not scalable to describe entire ecosystem rates.  Bottle-based methods may also only reliably estimate production and respiration associated with planktonic processes, whereas significant contributions to ecosystem metabolism can arise from other habitats, such as the benthos or intertidal marsh areas.  By contrast, open-water techniques have been increasingly used to estimate whole system metabolism given the availability of long-term, continuous time series of dissolved oxygen \citep{Odum56,Davanzo96}. Daily integrated measurements of metabolism represent the balance between daytime production and nighttime respiration attributed to all ecosystem components.  Open-water estimates also provide a basis for tracking ecosystem change over time and are more practical for capturing events or evaluating trends as compared to bottle-based approaches. Although metabolic rates vary naturally at different spatial and temporal scales \citep{Ziegler98,Caffrey04,Russell07}, anthropogenic nutrient sources are often contributing factors that increase rates of production (\citealt{Nixon95}, \citetalias{NRC00}).  Reliables estimates of whole ecosystem metabolism are critical for measuring both background rates of production and potential impacts of human activities on ecosytem condition.

The ability to accurately estimate whole system metabolism using the open-water method depends on the degree to which assumptions of the theory are met \citep{Staehr10,Kemp12}.  The fundamental assumption is that the time series of \ac{DO} represents a Lagrangian specification of the flow field that describes the same water mass over time \citep{Needoba12}.  The Lagrangian specification assumes that the time series characterizes individual fluid parcels regardless of location, as in a parcel of water moving with the tide.  In reality, most \ac{DO} time series are collected at fixed locations such as a mooring or dock, which is characteristized by an Eulerian specification of the flow field.  Time series at fixed locations may characterize water masses with different metabolic histories if water particles are transported by physical advection. A Lagrangian flow field is often assumed, such that estimates of metabolism may be inaccurate if substantial variation in water column mixing occurs throughout the period of observation \citep{Kemp80,Russell07}.  Given this critical challenge, the open-water method has been used with varying success in estuaries influenced by tidal mixing \citep{Caffrey04,Russell07,Caffrey14}.  In contrast, the method has been more successfully applied to water quality time series in lakes, although stratification may limit estimates to specific vertical layers \citep{Staehr10,Coloso11,Batt12}. Appropriate placement of monitoring sondes, sampling frequency and duration, and reliability of data from single stations have been relevant issues in applying the open-water method to systems influenced by physical mixing \citep{Russell07,Staehr10}.  Individual sampling stations near bay inlets or along major tidal axes may produce \ac{DO} time series that fail to meet the assumptions of the open-water method.

Although numerous studies have shown that application of the open-water method to lakes or estuaries may be problematic \citep{Ziegler98,Caffrey03,Coloso11,Batt12,Nidzieko14}, very few quantitative approaches have been developed to address potential bias or noise in \ac{DO} signals from physical advection.  For example, an extensive analysis by \citet{Caffrey03} applied the open-water method to estimate metabolism at 28 continuous monitoring stations at 14 US estuaries.  A significant portion of the production and respiration estimates were negative (3 - 69\% depending on site), suggesting advection of water masses was a likely factor influencing the \ac{DO} time series.  These `anomalous' values are typically omitted from the analysis \citep{Caffrey03,Collins13}, which may upwardly bias estimates of metabolism \citep{Caffrey14}.  Further, \citet{Nidzieko14} evaluated the effects of tidal advection on metabolism estimates in a mesotidal estuary.  Estimates from a single location were strongly correlated with the spring-neap cycle such that net heterotrophy was more common during spring tides, whereas metabolism was generally balanced during neap tides.  A control-volume approach was used by impounding a section of the upper estuary to understand how physical processes contribute to biological variability.  Although useful as an \textit{in situ}, site-specific approach, more accessible statistical methods specific to time series are needed given the increasing availability of continuous monitoring data. For example, \citet{Batt12} explored the use of a Kalman filter \citep{Harvey89} to remove process and observation uncertainty from \ac{DO} time series in lakes.  Similar approaches have not been developed for estuaries, particularly those that address potential effects of tidal advection.

This article describes the development and application of a method for improving estimates of ecosytem metabolism computed from \ac{DO} time series.  Specifically, the apparent effects of tidal advection on \ac{DO} observations are removed to improve the fidelity of open-water metabolism estimates derived from continuous water quality data.  We used a weighted regression approach originally developed to resolve trends in pollutant concentrations in streams and rivers \citep{Hirsch10}.  The weighted regression approach creates dynamic predictions of \ac{DO} as a function of time and tidal height change, which are then used to filter, or detide, the \ac{DO} signal.  The model is based on the recognition that daily fluctuations in \ac{DO} are caused by metabolism associated with the solar cycle, whereas other fluctuations in estuaries are likely associated with water level changes that generally exhibit pregression relative to the solar cycle.  The weighted regression model was applied, rather than methods commonly used for detiding in physical oceanography, to allow for the complex and dynamic patterns of \ac{DO} changes relative to advection.  First, we used simulated \ac{DO} time series with known characteristics to evaluate ability of the weighted regression to remove the simulated effects of a tidally-advected \ac{DO} gradient.  Second, the simulation results informed the application of the method to four case studies chosen from the \aclu{NERRS} (\acs{NERRS}, \citealt{Wenner04}).  In all examples, tidal height is used as a proxy for lateral water movements that may influence \ac{DO} observations.  In the absence of quantitative data describing lateral \ac{DO} variation (e.g., contemporaneous stations along a tidal axis), we assume tidal height is an appropriate characterization of lateral variation.  Accordingly, `tidal variation' or `changes in tidal height' are used throughout in reference to assumed lateral \ac{DO} gradients that are carried past monitoring sensors by tidal currents.

>>>>

Ecosystem metabolism describes the balance between production and respiration processes that create and consume organic matter \citep{Kemp12,Needoba12}.  Light exposure experiments of water samples collected at discrete locations and times have traditionally been used to measure metabolic activity.  Although highly controlled and precise, bottle-based methods are labor-intensive and not scalable to describe entire ecosystem rates.  Bottle-based methods may also only reliably estimate production and respiration associated with planktonic processes, whereas significant contributions to ecosystem metabolism can arise from other habitats, such as the benthos or intertidal marshes.  By contrast, open-water techniques have been increasingly used to estimate whole system metabolism given the availability of long-term, continuous time series of dissolved oxygen \citep{Odum56,Davanzo96}. Daily integrated measurements of metabolism represent the balance between daytime production and nighttime respiration attributed to all ecosystem components.  Open-water estimates also provide a basis for tracking ecosystem change over time and are more effective for capturing events or evaluating trends as compared to bottle-based approaches. Although metabolic rates vary naturally at different spatial and temporal scales \citep{Ziegler98,Caffrey04,Russell07}, anthropogenic nutrient sources are often contributing factors that increase rates of production (\citealt{Nixon95}, \citetalias{NRC00}).  Reliables estimates of whole ecosystem metabolism are critical for measuring both background rates of production and potential impacts of human activities on ecosytem function.

The ability to accurately estimate whole system metabolism using the open-water method depends on the degree to which assumptions of the theory are satisfied \citep{Staehr10,Kemp12}.  The fundamental assumption is that the time series of \ac{DO} represents a Lagrangian specification of the flow field \citep{Needoba12}.  The Lagrangian specification characterizes an individual fluid parcel through time, such as a parcel of water moving with the tide.  In reality, most \ac{DO} time series are collected at fixed locations such as a mooring or dock, so they are a Eulerian specification of the flow field.  Time series at fixed locations may characterize water masses with different metabolic histories as water is advected past the sensor, leading to errors in metabolism estimates \citep{Kemp80,Russell07}.  Given this critical challenge, the open-water method has been used with varying success in estuaries influenced by tidal transport and mixing \citep{Caffrey04,Russell07,Caffrey14}.  In contrast, the method has been more successfully applied to water quality time series in lakes, although stratification may limit estimates to specific vertical layers \citep{Staehr10,Coloso11,Batt12}. Appropriate placement of monitoring sondes, sampling frequency and duration, and reliability of data from single stations have been relevant issues in applying the open-water method to systems influenced by physical mixing \citep{Russell07,Staehr10}.  Individual sampling stations near bay inlets or along major tidal axes may produce \ac{DO} time series that are strongly influenced by advection leading to relatively large errors when using the open-water method.

Although numerous studies have shown that application of the open-water method to lakes or estuaries may be problematic \citep{Ziegler98,Caffrey03,Coloso11,Batt12,Nidzieko14}, very few quantitative approaches have been developed to address potential bias or noise in metabolism estimates resulting from advection.  For example, an extensive analysis by \citet{Caffrey03} applied the open-water method to estimate metabolism at 28 continuous monitoring stations at 14 US estuaries.  A significant portion of the production and respiration estimates were negative (3 - 69\% depending on site), suggesting advection of water masses was a likely a major factor influencing the \ac{DO} time series.  These `anomalous' values are typically omitted from the analysis \citep{Caffrey03,Collins13}, which may upwardly bias estimates of mean metabolism \citep{Caffrey14}.  Further, \citet{Nidzieko14} evaluated the effects of tidal advection on metabolism estimates in a mesotidal estuary.  Estimates from a single location were strongly correlated with the spring-neap cycle such that net heterotrophy was more common during spring tides, whereas metabolism was generally balanced during neap tides.  This examples illustrates that metabolism may actually be modulated by the tide, as opposed to artifacts in the \ac{DO} time series caused by advection on daily time scales.  \citet{Nidzieko14} used a control-volume approach by impounding a section of the upper estuary to understand how physical processes contribute to biological variability.  Although useful as an \textit{in situ}, site-specific approach, more accessible statistical methods specific to time series are needed given the increasing availability of continuous monitoring data. For example, \citet{Batt12} explored the use of a Kalman filter \citep{Harvey89} to remove process and observation uncertainty from \ac{DO} time series in lakes.  Similar approaches have not been developed for estuaries, particularly those that address potential effects of tidal advection.

This article describes the development and application of a method for improving estimates of ecosytem metabolism computed from \ac{DO} time series by reducing the effect of tidal advection on continuous \ac{DO} observations.  We used a weighted regression approach originally developed to resolve trends in pollutant concentrations in streams and rivers \citep{Hirsch10}.  The weighted regression approach creates dynamic predictions of \ac{DO} as a function of time and tidal height change, which are then used to filter, or detide, the \ac{DO} signal.  The model is based on the recognition that daily fluctuations in \ac{DO} caused by metabolism are associated with the solar cycle, whereas other fluctuations in estuaries are likely associated with cyclical water level changes that generally exhibit pregression relative to the solar cycle.  The weighted regression model was applied, rather than methods commonly used for detiding in physical oceanography, to allow for the complex and dynamic patterns of \ac{DO} changes relative to advection.  During initial method development, we used simulated \ac{DO} time series with known characteristics to evaluate ability of the weighted regression to remove the simulated effects of a tidally-advected \ac{DO} gradient.  Subsequently, the simulation results informed further development of the method as applied to four case studies chosen from the \aclu{NERRS} (\acs{NERRS}, \citealt{Wenner04}).  In all examples, tidal height is used as a proxy for advection because it was recorded with the \ac{NERRS} data.  In the absence of quantitative data describing lateral \ac{DO} variation (e.g., contemporaneous stations along a tidal axis), the model determines the advective effect empirically.

### 2015-02-03 titlepage.tex
almost done with Smith revisions
finished revisions from internal review
finished revisions from Hagy comments after Smith
cover letter, some edits from Jim, waiting on final comments before first submissions

\begin{singlespace}
\title{{\bf {\Large Evaluation and correction of noise related to physical processes in estimates of estuary metabolism}}}
\author{
  {\bf {\normalsize Marcus W. Beck$^1$, Michael C. Murrell$^2$, James D. Hagy III$^2$}}
  \\\\{\textit {\normalsize $^1$ORISE Research Participation Program}}
  \\{\textit {\normalsize USEPA National Health and Environmental Effects Research Laboratory}}
	\\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
	\\{\textit {\normalsize Phone: 850-934-2480, Fax: 850-934-2401, Email: \href{mailto:beck.marcus@epa.gov}{beck.marcus@epa.gov}}}
  \\\\{\textit {\normalsize $^2$USEPA National Health and Environmental Effects Research Laboratory}}
	\\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
	\\{\textit {\normalsize Phone: 850-934-2433, Fax: 850-934-2401, Email: \href{mailto:murrell.michael@epa.gov}{murrell.michael@epa.gov}}}
  \\\\{\textit {\normalsize $^3$USEPA National Health and Environmental Effects Research Laboratory}}
	\\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
	\\{\textit {\normalsize Phone: 850-934-2455, Fax: 850-934-2401, Email: \href{mailto:hagy.jim@epa.gov}{hagy.jim@epa.gov}}}
	}
\date{}
\maketitle
\vfill{\centerline{\textit {\normalsize Running head: Noise in Estuary Metabolism}}}
\end{singlespace}
\clearpage

>>>>

\begin{singlespace}
\title{{\bf {\Large Improving estimates of ecosystem metabolism by reducing effects of tidal advection on dissolved oxygen time series}}}
\author{
  {\bf {\normalsize Marcus W. Beck$^1$, James D. Hagy III$^2$, Michael C. Murrell$^3$}}
  \\\\{\textit {\normalsize $^1$ORISE Research Participation Program}}
  \\{\textit {\normalsize USEPA National Health and Environmental Effects Research Laboratory}}
	\\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
	\\{\textit {\normalsize Phone: 850-934-2480, Fax: 850-934-2401, Email: \href{mailto:beck.marcus@epa.gov}{beck.marcus@epa.gov}}}
  \\\\{\textit {\normalsize $^2$USEPA National Health and Environmental Effects Research Laboratory}}
  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
	\\{\textit {\normalsize Phone: 850-934-2455, Fax: 850-934-2401, Email: \href{mailto:hagy.jim@epa.gov}{hagy.jim@epa.gov}}}
  \\\\{\textit {\normalsize $^3$USEPA National Health and Environmental Effects Research Laboratory}}
	\\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
	\\{\textit {\normalsize Phone: 850-934-2433, Fax: 850-934-2401, Email: \href{mailto:murrell.michael@epa.gov}{murrell.michael@epa.gov}}}
	}
\date{}
\maketitle
\vfill
{\centerline{\it {\normalsize Key words: Dissolved Oxygen, Ecosystem Metabolism, Open Water Method, Tides,}}}
{\centerline{\it {\normalsize Time Series, Weighted Regression}}}
{\centerline{\it {\normalsize }}}
{\centerline{\it {\normalsize Running head: Improving Estimates of Estuary Metabolism}}}
\end{singlespace}
\clearpage
